# Final Project 
# EDA on the Chess Games dataset
## By: Abhijeet Vichare
## Date: 12/05/2022

## Table of Contents
1. Introduction
2. Data
3. Data Cleaning
4. Exploratory Data Analysis
5. Model Building
6. Conclusion

1. Data Wrangling
    a. Data importing
    b. Checking the data types of variables and performing necessary steps to change the data type based on relevancy.
    c. Univariate Statistics of the variables
    d. Dealing with null values
    e. Outlier analysis
2. Data Visualization
    a. Bivariate analysis
    b. Correlation analysis
3. Feature
    a. Chi square test
    b. ANOVA test
    c. Correlation test
4. Linear Regression
    a. Linear Regression assumption validation
    b. Model training
    c. Hyperparameter tuning
    d. Regularization
    e. Result visualization
    
## 1. Introduction

We have used the [chess dataset]{https://www.kaggle.com/datasets/datasnaek/chess} from kaggle which 
contains the data of chess games played by different players. The dataset contains 20058 rows and 16 columns
and its collected from Lichess.org. A short description of the dataset: 

1. id: Unique ID for the game 
2. rated: Indicates whether the game was rated or unrated
3. created_at: Timestamp of the game creation
4. last_move_at: Timestamp of the last move
5. turns: Number of turns in the game
6. victory_status: Indicates the status of the game (win, draw, outoftime, resign, mate)
7. winner: Indicates the winner of the game (white, black, draw)
8. increment_code: Increment code
9. white_id: Unique ID of the white player
10. white_rating: Rating of the white player
11. black_id: Unique ID of the black player
12. black_rating: Rating of the black player
13. moves: Moves in the game
14. opening_eco: Opening ECO
15. opening_name: Opening name
16. opening_ply: Opening ply

## 2. Data
### 2.1 Importing the libraries
```{r}
# install.packages(c("assertive.data","plotly","tidyverse", "ggplot2", "dplyr", "ggcorrplot", "corrplot", "caret", "rpart", "rpart.plot", "rattle", "rpart.plot"))
```
```{r}
library(ggplot2)
library(corrplot)
library(caret)
library(plotly)
library(assertive.data)
source("model_training.R")
```

### 2.2 Importing the dataset
```{r}
chess <- read.csv("games.csv")
```
```{r}
colnames(chess)
```
### 2.3 Data Exploration
```{r}
#types of data
str(chess)
#summary of data
summary(chess)
#checking for null values
colSums(is.na(chess))
```
There are 20058 rows and 16 columns in the dataset. Columns like rated can be converted to boolean. 
We can convert the numerical created_at and last_move_at to date time format. The columns like white_id, black_id can be dropped as they are not relevant to the analysis. The column id can be dropped as it is just a unique identifier for the game. The column turns can be dropped as it is highly correlated with the column white_rating. The column victory_status can be dropped as it is highly correlated with the column winner.
There are no null values in the dataset.
```{r}
# converting rated to boolean
chess$rated <- as.logical(chess$rated)
# converting created_at and last_move_at to date time format
chess$created_at <- as.POSIXct(chess$created_at/1000, origin = "1970-01-01")
chess$last_move_at <- as.POSIXct(chess$last_move_at/1000, origin = "1970-01-01")
```
```{r}
# difference of created_at and last_move_at
chess$match_length <- chess$last_move_at - chess$created_at
# limit the max to 2000s
chess$match_length <- ifelse(chess$match_length > 1500, 1500, chess$match_length)
```


### 2.4 Univariate Visualization
```{r}
# convert match_length to float and plot the histogram
chess$match_length <- as.numeric(chess$match_length)
ggplot(chess, aes(x = match_length)) + geom_histogram(bins = 100)
```
There are outliers at 0 and 1500. We can remove the outliers and plot the histogram again.
```{r}
# remove the outliers and plot the histogram
temp <- filter(chess, match_length > 0 & match_length < 1500)
ggplot(temp, aes(x = match_length)) + geom_histogram(bins = 100)
```

```{r}
#histogram of white_rating
ggplot(chess, aes(x = white_rating)) + geom_histogram(bins = 100)
```
```{r}
#histogram of black_rating
ggplot(chess, aes(x = black_rating)) + geom_histogram(bins = 100)
```
```{r}
# combine the white_rating and black_rating into one vector and plot the histogram
rating <- c(chess$white_rating, chess$black_rating)
ggplot(data.frame(rating), aes(x = rating)) + geom_histogram(bins = 100)
```
```{r}
#value counts of rating and sort it in descending order

rating <- c(chess$white_rating, chess$black_rating)
rating <- as.data.frame(table(rating))
rating <- rating[order(-rating$Freq),]
head(rating)
```
The ratings follow a normal distribution. The mean of the ratings is 1591. There is one outlier of  1500 where The most common rating is 1500.

```{r}
# Distribution of the Victory Status
ggplot(chess, aes(x = victory_status)) + geom_bar()
#value counts as percentage of victory_status and sort it in descending order
victory_status <- as.data.frame(table(chess$victory_status))
victory_status$percent <- round(victory_status$Freq/sum(victory_status$Freq)*100, 2)
victory_status <- victory_status[order(-victory_status$percent),]
victory_status
```
55.5% of the games are resigned and 31.53% of the games end by a mate.
```{r}
# Turn Distribution
ggplot(chess, aes(x = turns)) + geom_histogram(bins = 100)
summary(chess$turns)
```
```{r}
# Distribution of the Winner
ggplot(chess, aes(x = winner)) + geom_bar()
#value counts as percentage of winner and sort it in descending order
winner <- as.data.frame(table(chess$winner))
winner$percent <- round(winner$Freq/sum(winner$Freq)*100, 2)
winner <- winner[order(-winner$percent),]
winner
```

```{r}
# Divide the increment code into two columns
chess$increment_code <- as.character(chess$increment_code)
# everthing before the plus sign
chess$increment_code1 <- strsplit(chess$increment_code, split = "\\+")
chess$increment_code1 <- sapply(chess$increment_code1, function(x) x[1])
# everything after the plus sign
chess$increment_code2 <- strsplit(chess$increment_code, split = "\\+")
chess$increment_code2 <- sapply(chess$increment_code2, function(x) x[2])
```
```{r}
# convert to numeric and plot the histogram
chess$increment_code1 <- as.numeric(chess$increment_code1)
ggplot(chess, aes(x = increment_code1)) + geom_histogram(bins = 100)
```
```{r}
# convert to numeric and plot the histogram
chess$increment_code2 <- as.numeric(chess$increment_code2)
ggplot(chess, aes(x = increment_code2)) + geom_histogram(bins = 100)
```
```{r}
# Median of the increment code1
median(chess$increment_code1)
# Median of the increment code2
median(chess$increment_code2)
```

The median base time of match is 10 minutes and the increment is 0 second which is sudden death.
```{r}
#value counts as percentage of opening_name and sort it in descending order
opening_name <- as.data.frame(table(chess$opening_name))
opening_name$percent <- round(opening_name$Freq/sum(opening_name$Freq)*100, 2)
opening_name <- opening_name[order(-opening_name$percent),]
head(opening_name)
dim(opening_name)
```
There are 1477 unique opening names. The most common opening name is Van't Kruijs Opening. 

### 2.5 Bivariate Visualization
Points to check: 
- Is there a relationship between the match length and the ratings?
- Is there a relationship between the match length and the victory status?
- Is there a relationship between the match length and the winner?
- Is there a relationship between the match length and the increment code?
- Is there a relationship between the match length and the opening name?
- Is there a relationship between the ratings and the victory status?
- Is there a relationship between the ratings and the winner?
- Is there a relationship between the ratings and the increment code?
- Is there a relationship between the ratings and the opening name?
- Is there a relationship between the victory status and the winner?
- Is there a relationship between the victory status and the increment code?
- Is there a relationship between the victory status and the opening name?
- Is there a relationship between the winner and the increment code?
- Is there a relationship between the winner and the opening name?
- Is there a relationship between the increment code and the opening name?
```{r}
# Is there a relationship between the match length and the ratings?
ggplot(chess, aes(x = match_length, y = white_rating)) + geom_point()
ggplot(chess, aes(x = match_length, y = black_rating)) + geom_point()
# correlation between match_length and white_rating and the best fit line
cor.test(chess$match_length, chess$white_rating)
# correlation between match_length and black_rating and the best fit line
cor.test(chess$match_length, chess$black_rating)
```
There seems to be no relationship between the match length and the ratings.
```{r}
# Is there a relationship between the match length and the victory status?
# Check the relationship between a catergorical variable and a numerical variable
ggplot(chess, aes(x = match_length, y = victory_status)) + geom_boxplot()
```
The resigns and mates have a lower match length than the others. 
The out_of_time has the highest match length than the others. Is it a correlation or a causation?
```{r}
# Is there a relationship between the match length and the winner?
ggplot(chess, aes(x = match_length, y = winner)) + geom_boxplot()
```
If it's a draw the match length is longer than the others. It seems natural as the draw matches tend
to be longer than the others towards the end of the game. 
```{r}
# Is there a relationship between the match length and the increment code?
# Bin the increment_code1 into bins based on custom bins
chess$increment_code1_bin <- cut(chess$increment_code1, breaks = c(0, 20, 50, 100, 200, 1000))
ggplot(chess, aes(x = match_length, y = increment_code1_bin)) + geom_boxplot()
```
The match length seems to be lower for increment codes between 0-20 as that seems natural due to the 
sudden death nature of the game. Surprisingly, the match length of increment codes between 50-100 is lower
than the match length of increment codes between 20-50. 

```{r}
# Is there a relationship between the ratings and the victory status?
ggplot(chess, aes(x = white_rating, y = victory_status)) + geom_boxplot()
ggplot(chess, aes(x = black_rating, y = victory_status)) + geom_boxplot()
```
There seems to be no relationship between the ratings and the victory status.
```{r}
# Is there a relationship between the ratings and the winner?
ggplot(chess, aes(x = white_rating, y = winner)) + geom_boxplot()
ggplot(chess, aes(x = black_rating, y = winner)) + geom_boxplot()
```
As expected, the higher the rating the higher the chance of winning. 
```{r}
# Let's check the difference in ratings between the winner and the loser
chess$rating_diff <- chess$white_rating - chess$black_rating
# Put the title of the plot as the difference in ratings
ggplot(chess, aes(x = rating_diff, y = winner)) + geom_boxplot() + labs(title = "Difference in ratings (White - Black) Vs Winner")
```
It can be seen in the boxplots that majority of the draws have a difference in ratings of 0. 
And the difference is postive when white wins and negative when black wins with a few outliers.

```{r}
# Is there a relationship between the ratings and the increment code?
# Bin the increment_code1 into bins based on custom bins
chess$increment_code1_bin <- cut(chess$increment_code1, breaks = c(0, 20, 50, 100, 200, 1000))
ggplot(chess, aes(x = white_rating, y = increment_code1_bin)) + geom_boxplot()
ggplot(chess, aes(x = black_rating, y = increment_code1_bin)) + geom_boxplot()
```
There seems to be no relationship between the ratings and the increment code.
```{r}
chess$increment_code2_bin <- cut(chess$increment_code2, breaks = c(0, 20, 50, 100, 200, 1000))
```
```{r}
# Is there a relationship between the victory status and the winner?
# Create a table with counts across the two variables
tab <- table(chess$victory_status, chess$winner)
tab
# Create a table with percentages across the two variables
tab_percent <- round(prop.table(tab, 1)*100, 2)
tab_percent
```
The table shows that black and white are equally probable to run out of time. There's a small advantage
for white to resign and black to mate. 
```{r}
# Is there a relationship between the victory status and the increment code?
# Create a table with counts across the two variables
tab <- table(chess$victory_status, chess$increment_code1)
tab
# Create a table with percentages across the two variables
tab_percent <- round(prop.table(tab, 1)*100, 2)
tab_percent
```

```{r}
# Is there a relationship between the winner and the increment code?
# Create a table with counts across the two variables
tab <- table(chess$winner, chess$increment_code1_bin)
tab

tab_percent <- round(prop.table(tab, 1)*100, 2)
tab_percent
```

```{r}
# Let's do some analysis on Moves
# Is there a rela tionship between the match length and the number of moves?
# correlation between match_length and moves and the best fit line
cor.test(chess$match_length, chess$turns)
```
Ther's seems a weak correlation between the match length and the number of moves. 
```{r}
# Deep down into the column moves
# Create a new column with the number of moves
chess$number_of_moves <- sapply(chess$moves, function(x) length(strsplit(x, " ")[[1]]))

# correlation between turns and number_of_moves and the best fit line
cor.test(chess$turns, chess$number_of_moves)
```
```{r}
# confirming that the number of moves and turns is equal for all rows.
chess[chess$number_of_moves != chess$turns]
```

# 3 Feature Engineering and Selection
    a. Chi square test
    b. ANOV A test
    c. Correlation test

Given the data, we are going to predict the winner of the game. We'll treat black and white as 1, and 0 respectively. 
We'll remove the draw cases as they are very few in number. 
```{r}
# Remove the draws
chess <- subset(chess, winner != "draw")
# Convert the winner column to 1 and 0
chess$winner <- ifelse(chess$winner == "white", 1, 0)
```

We'll perform chi square test to check the relationship between the categorical variables and the winner.
```{r}
# get column names
colnames(chess)
```

# Chi square test
# Catergorical variables that we will consider:
# victory_status, increment_code1_bin, rated
# Create a function to perform chi square test
```{r}
chisq_test <- function(x, y) {
  # Create a table with counts across the two variables
  tab <- table(x, y)
  # Create a table with percentages across the two variables
  tab_percent <- round(prop.table(tab, 1)*100, 2)
  # Perform chi square test
  chi_sq <- chisq.test(tab)
  # Print the results
  print(chi_sq)
  print(tab_percent)
}
```
```{r}
# Loop over categorical variables and perform chi square test
for (i in c("victory_status", "increment_code1_bin", "increment_code2_bin", "rated")) {
  print(i)
  chisq_test(chess[[i]], chess$winner)
}
```
```{r}
# chisq_test(chess$increment_code1_bin, chess$winner)
head(chess$increment_code1_bin)
```


# ANOVA test
# Numerical variables that we will consider:
# white_rating, black_rating, rating_diff, match_length, turns, number_of_moves,
# increment_code1, increment_code2
# Create a function to perform ANOVA test
```{r}
anova_test <- function(x, y) {
  # Perform ANOVA test
  anova <- aov(x ~ y)
  # Print the results
  print(anova)
#   get the summary
  print(summary(anova))
}
```
```{r}
# Loop over numerical variables and perform ANOVA test
for (i in c("white_rating", "black_rating", "rating_diff", "match_length",  "increment_code1", "increment_code2")) {
  print(i)
  anova_test(chess[[i]], chess$winner)
}
```
Increment codes have no relationship with the winner. black_rating and white_rating have a high
significance with the winner. Rating diff also has a high significance with the winner. 


```{r}
# Correlation test
# Numerical variables that we will consider:
# white_rating, black_rating, rating_diff,

# Create a function to perform correlation test
cor_test <- function(x, y) {
  # Perform correlation test
  cor <- cor.test(x, y)
  # Print the results
  print(cor)
}
```
```{r}
# Loop over numerical variables and perform correlation test
for (i in c("white_rating", "black_rating", "rating_diff")) {
  print(i)
  cor_test(chess[[i]], chess$winner)
}
```
```{r}
# correaltion between the numerical variables
cor(chess[, c("white_rating", "black_rating", "rating_diff")])
```

# 4. Linear Regression
a. Linear Regression assumption validation

```{r}
source("model_training.R")
# Preprocessing function
chess <- read.csv("games.csv")
# Preprocess the data
chess <- preprocess(chess)
chess <- chess %>% select(-rating_diff) # nolint
model <- train_model(chess)
```
```{r}
```
```{r}
# Linear Regression assumption validation
# 1. Linearity
# Let's check the linearity of the numerical variables
# Create a function to check the linearity of the numerical variables
linearity_test <- function(x, y) {
  # Perform linear regression
  lm <- lm(x ~ y)
  # Print the results
  print(lm)
  print(summary(lm))
}
```

```{r}
# Loop over numerical variables and perform linear regression
for (i in c("white_rating", "black_rating")) {
  print(i)
  linearity_test(chess[[i]], chess$winner)
}
```
```{r}
# 2. Normality
# Let's check the normality of the numerical variables
# Create a function to check the normality of the numerical variables
normality_test <- function(x) {
  # Perform Shapiro-Wilk test
    shapiro <-  shapiro.test(x)
    print(shapiro)
}
```
```{r}
# Loop over numerical variables and perform Shapiro-Wilk test
for (i in c("white_rating", "black_rating")) {
  print(i)
#   sample 5000 rows and perform the test
    normality_test(sample(chess[[i]], 5000))
}
```

b. Model training
```{r}
# Model training
train_model <- function(data,verbose = FALSE) {
  # Split the data into train and test
  train_index <- createDataPartition(data$winner, p = 0.7, list = FALSE)
  train <- data[train_index, ]
  test <- data[-train_index, ]
  # Train the model
  model <- glm(winner ~ ., data = train, family = "binomial")
  # Predict the test data
  pred <- predict(model, test)
  if (verbose) {
    print(model)
    print(summary(model))
    # Print the confusion matrix
    print(table(test$winner, pred > 0.5))
    # Print the accuracy
    print(mean(test$winner == (pred > 0.5)))

  }
  return(model)
}
```



